Predicting I/O-performance in HPC using Artificial Neural Networks

### 2016-06-25 Paper/template.tex
analysis

\usepackage{amssymb}
\usepackage{float}
\usepackage[position=t,singlelinecheck=off]{subfig}
\usepackage{caption}
% ------------

@@

\section{Structure of the analysis}
The storage system can be analyzed with series of measurements.
We use synthetic benchmark-tests were many similar measurements follow each other, instead of using arbitrary access parameters for consecutive accesses.
This approach allows for a in-depth analysis of system behavior for different use cases.\\
The file access parameters are:
\begin{itemize}
	\item \textbf{File-ID}: Identifies the invoked file.
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{OpType}: Type of file access (read or write).
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time in seconds from initiation of the file access to the termination of the process.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed is not directly available.

@@

\subsubsection{I/O-path for access time prediction}
Due to the exponential decay of processing speed in the hierarchy, file accesses with similar access parameters, but varying I/O-path, are differentiated with a step in the magnitude of access time.
As the access time is dominated by the slowest component along the I/O-path, a step in the measured access time occurs between to I/O-paths of diverging length.
In figure \ref{fig:io_paths} measurements of reading file accesses can seen, this figure will be discussed in further detail in the following chapter.
\begin{figure}
	\centering
	\includegraphics[width=.47\textwidth]{src/plot_SizeSorted_log_read_seq.png}
	\caption{Measurements of reading file accesses. Access sizes increase from left to right and all measurements with equal access parameter values have the same color.}
	\label{fig:io_paths}
\end{figure} 
The different groups of measurements with equal access parameter values (points with the same color) are clearly visible.
Like described previously the different groups are differentiated with a step in the magnitude of access time and can be explained with varying I/O-paths.

\subsection{Error classes}

\subsection{Models}

\section{Evaluation}

@@

\subsection{Prediction of file accesses}

>>>>

\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{subcaption}
% ------------

@@

\section{Structure of the analysis}
The storage system can be analyzed with series of measurements.
We use synthetic benchmark-tests were many similar measurements follow each other, instead of using arbitrary access parameters for consecutive accesses.
This approach allows for a in-depth analysis of system behavior for different use cases.\\
The file access parameters are:
\begin{itemize}
	\item \textbf{File-ID}: Identifies the invoked file.
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{Access type}: Differentiates reading and writing file accesses.
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time in seconds from initiation of the file access to the termination of the process.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed is not directly available.

@@

\subsubsection{I/O-path for access time prediction}
Due to the exponential decay of processing speed in the hierarchy, file accesses with similar access parameters, but varying I/O-path, are differentiated with a step in the magnitude of access time.
As the access time is dominated by the slowest component along the I/O-path, a step in the measured access time occurs between to I/O-paths of diverging length.
In figure \ref{src/plot_SizeSorted_log_read_seq.png} measurements of reading file accesses can seen, this figure will be discussed in further detail in the following chapter.
\fig{width=.47\textwidth}{src/plot_SizeSorted_log_read_seq.png}{Measurements of reading file accesses. Access sizes increase from left to right and all measurements with equal access parameter values have the same color.}\\
The different groups of measurements with equal access parameter values (points with the same color) are clearly visible.
Like described previously the different groups are differentiated with a step in the magnitude of access time and can be explained with varying I/O-paths.

\subsection{Error classes}
\label{sec:error_classes}
Error classes are supposed to approximate I/O-paths using the available measurements of file access times.
To derive such knowledge we need a characteristic that is typical for the different I/O-paths. Knowing the file access parameters and the approximation of I/O-path should allow us to predict the magnitude of access time for this process.\\
At a first thought one could come up with the idea of clustering the data seen in figure \ref{fig:io_paths}, because the groups of measurements with varying I/O-paths might be correctly differentiated by it. 
This, however, is not a good approach, one would have to cluster for every set of access parameters individually, otherwise the clusters would be meaningless. In a real application scenario a large number of different access parameters with only very few instances occur, which makes this approach unfeasible.\\
Another idea might be then to use the difference of an arbitrary value to the measured file access time as parameter for the clustering. 
Such a value would not be appropriate for all magnitudes of access time occurring in the data, I/O-paths differentiating measurements with small access time might not be discriminated, because the gap between them doesn't matter at the scale of the chosen arbitrary value.\\
Our approach uses the residues of a model like linear regression that can't differentiate between measurements with equal access parameter values.
Basically the linear regression predicts an average access time for these measurements.
The difference of an individual measurement for these access parameters to the average value is then characteristic for the I/O-path.
Clustering the residues with a k-Means algorithm allows us then to assign each measurement to a specific cluster, which is the approximation of I/O-path.
We call the retrieved clusters \textbf{error classes}.
We use the absolute error of linear regression as clustering parameter so that one cluster can for example represent an deviation of 1 millisecond.
Ideally it might be the case that file accesses that took 1 millisecond longer than usual were processed on the same I/O-path.
In that case error classes are directly corresponding to I/O-paths. \medskip

We will use error classes to quantify the importance of knowledge about I/O-paths for access time prediction.
For the actual prediction error classes can't be used, because a measured access time is required to determine the error class of a measurement.
However, the method could be used for a tool, which analyses the I/O made during application of a program. 
It could analyze whether more file accesses throughout execution were slower than expected and therefore used unfavorable I/O-paths.
If a direct correlation of error class to I/O-paths were found, such an tool might be able to inform about percentages of used I/O-paths.

\subsection{Models}
Different models were used to predict access times of file accesses.
Each model seeks to correlate its known file access parameters to the corresponding access time. 
Linear regression was used as a baseline model with a simple mapping of access size to access time.
Additionally, three models with different input information utilizing ANNs were used:
\begin{itemize}
	\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
	\item A second ANN-model, called \textbf{ema-model}, received additionally to the parameters of the previous model information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance.
	\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
\end{itemize}\medskip

The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.\\
Figure \ref{src/plot_From100001to100250_read_seq.png} shows a small series of measurements of sequentially read file accesses.
The peaks in access time are occurring periodically with knowledge about the period of peaks better access time predictions could be made.
\fig{width=.47\textwidth}{src/plot_From100001to100250_read_seq.png}{An example of a figure}\\
Our second model is supposed to use its input information of previous data throughput for that.
The data throughput has peaks the same period as the access time.
Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
\begin{equation}
EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
\end{equation}
With $t(i)$ the access time of measurement $i$.
The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
In figure \ref{src/ema2.png} the EMA-function to a graph with periodic peaks can be seen.
\fig{width=.47\textwidth}{src/ema2.png}{Exponential moving average in red for a graph with perdiodic peaks.}\\
The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions.\medskip

The error-class-model is used as described in chapter \ref{sec:error_classes} to examine the potential improvements for access time prediction due to knowledge about I/O-paths.
\section{Evaluation}

@@

\subsection{Prediction of file accesses}
From the residues of our models (see table \ref{residues}) it became clear that access times shouldn't be modeled linearly.
Furthermore, it has been found to be difficult to exploit time dependencies of the storage system.
As our hypothesis predicted, \textbf{it is essential to use knowledge about I/O-paths for precise predictions of access times}.
The error classes allow the ANN to decrease the error significantly.
In figure \ref{preds} the benefit of error classes as input can be seen:
The second ANN is often able to classify file accesses correctly in respect to their I/O-path.
